Interpolation of Sparse High-Dimensional Data\thanks {This work was supported by the National Science Foundation Grant CNS-1565314.}

### 2020-08-13 [2019-02]_NUMA_(interpolant_error_bound)/tchlux-NA-interpolation/Versions/0-paper.tex
Finished NUMA paper and submitted. Might need to do one more round with editors.

% Insert here the call for the packages your document requires
\usepackage{graphicx}
\usepackage{tikz}     % For adding axes to plots
\usetikzlibrary{positioning}
\usepackage{graphicx} % For including PDFs
\usepackage{amsmath}  % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{booktabs} % For formal tables
\usepackage{tabularx} % For tables with nice column widths.
\usepackage[colorlinks,allcolors=blue]{hyperref}
\usepackage{float}    % For controlling the placement of tables.
\usepackage{enumitem} % For changing item indents in enumerate.

%% -------------------------------------------------------------------
%% Command for changing the color of text within a block.
\usepackage{xcolor}   % Color package for coloring text.
\usepackage[normalem]{ulem} % For strikethrough text and the \sout{} command.
\usepackage[xcolor,rightbars]{changebar}% Package for adding bars into margin next to comments.
\renewcommand{\changebarsep}{5mm}   % Push the bar farther into the margin.
\renewcommand{\changebarwidth}{1mm} % Increase the width of the bar.
\cbcolor{red}
%% \newcommand{\changed}[2]{\cbstart{\color{red}#1}{\color{gray}\sout{#2}}\cbend{}}
\newcommand{\changed}[2]{\cbstart{\color{red}#1}{}\cbend{}}
%% -------------------------------------------------------------------

@@

%% Include the response to reviewers at the beginning of the document.
\include{tchlux-NA-response-2020-04}

%% Set the title.
\title{Interpolation of Sparse High-Dimensional Data\thanks
  {This work was supported by the National Science Foundation Grant CNS-1565314.}
}
% \subtitle{Do you have a subtitle?\\ If so, write it here}
%\titlerunning{Short form of title}        % if too long for running head

@@

\date{Received: date / Accepted: date}
%% The correct dates will be entered by the editor

{\let\newpage\relax\maketitle}

%% \maketitle

\setcounter{page}{1}

@@

Regression and interpolation have a considerable theoretical base in
one dimension \cite{cheney2009course}. \changed{Common techniques
  include Lagrangian interpolation, piecewise linear interpolation,
  and more generally spline interpolation via B-splines
  \cite{de1978practical}}{Splines in particular are well understood
  as an interpolation technique in one dimension
  \cite{de1978practical}, particularly B-splines}. Tensor products of
B-splines \cite{unther1996interpolating} or other basis functions have
an unfortunate exponential scaling with increasing
dimension. Exponential scaling prohibits tensor products from being
reasonably applied beyond three-dimensional data. In order to address
this dimensional scaling challenge, C. de Boor and others proposed box
splines \cite{de2013box}, of which one of the approximation techniques
in this work is composed \cite{lux2018novel}.

The theoretical foundation of low dimensional interpolation allows the
construction of strong error bounds that are \changed{lacking
  in}{absent from} high dimensional problems. \changed{A large body of
  literature exists studying the construction and application of
  sparse grids \cite{bungartz2004sparse,chkifa2014high}, which attempt
  to address the problem of dimensional scaling. The sparse grid point
  sets combined with specific approximation functions are used to
  construct total error bounds by considering projections onto
  orthogonal polynomials
  \cite{migliorati2013approximation,nobile2016convergence}. However,
  sparse grids have exponential size in dimension making them
  intractable for more than tens of dimensions and do not immediately
  apply to the problem of scattered data interpolation.}{} This work
extends some known results regarding the secant method
\cite{dennis1996numerical} to construct \changed{a uniform error bound
  for interpolants over arbitrary scattered data in any dimension}{an
  interpolation error bound for problems of arbitrary
  dimension}. These error bounds are useful, considering
\changed{similar uniform bounds}{the same cannot} cannot be
\changed{constructed}{said} for regression algorithms in general. The
maximum complexity of an interpolant is bounded by the amount of data
available, while the maximum complexity of a regressor is bounded by
both the amount of data and the chosen parametric form. For this
reason, generic uniform bounds are largely unobtainable for regression
techniques on arbitrary approximation problems, even when the
approximation domain is bounded.

@@

In this work, multivariate interpolation is defined given a closed
convex subset $Y$ of a metrizable topological vector space with metric
$s$, some function $f:\mathbb{R}^d \rightarrow Y$ and a set of points
$X = \bigl\{x^{(1)}$, $\ldots$, $x^{(n)}\bigr\} \subset \mathbb{R}^d$,
along with associated function values $f\bigl(x^{(i)}\bigr)$. The goal
is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow Y$
such that $\hat f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all
$i = 1$, $\ldots$, $n$. It is often the case that the form of the true
underlying function $f$ is unknown, however it is still desirable to
construct an approximation $\hat f$ with small approximation error at
$y \notin X$. The two metric spaces that will be discussed in this
work are the real numbers with metric $s(x,y) = |x-y|$, and the set of
cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov
(KS) statistic \cite{lilliefors1967kolmogorov} as a metric.

@@

The most challenging problem when scaling in dimension is that the
number of possible interactions between dimensions grows
exponentially. Quantifying all possible interactions becomes
intractable, and hence \changed{problems}{} beyond \changed{ten to
  twenty dimensions usually rely on linear models}{three-dimensional
  data mostly linear models are used}. That is not to say nonlinear
models are absent, but nonlinearities are often either preconceived or
model \changed{interactions between small subsets of}{pairwise
  interactions between} dimensions at most. Even globally nonlinear
approximations such as neural networks are constructed from
compositions of summed low-interaction functions
\cite{clevert2015fast}.

Provided the theoretical and practical motivations for exploring
interpolants, the main objective of this work is to study the
empirical performance differences in both accuracy and computational
expense between a set of interpolation techniques and a set of common
regression techniques. The main contributions of this work are a sharp
theoretical error bound for (piecewise) linear interpolation in
arbitrary dimension and multiple empirical comparisons that illustrate
potential advantages of interpolation.

The remainder of this paper is organized as follows. Sections
\ref{sec:regression} and \ref{sec:interpolation} present the
multivariate models. Section \ref{sec:error} gives the error measuring
methodology that is used for collecting empirical results. Section
\ref{sec:theory} contains the theoretical error bounds for
interpolation as well as an analytic test for demonstration. Section
\ref{sec:data} presents the data sets for empirical approximation
analysis and the approximation results for all models on these data
sets. Section \ref{sec:discussion} analyzes and discusses the results,
their implications, and their limitations. Finally, Section
\ref{sec:conclusion} concludes.

@@

\subsection{Multivariate Adaptive Regression Splines}
\label{sec:mars}
This approximation was introduced in \cite{friedman1991multivariate}
and subsequently improved to its current version in
\cite{stanford1993fast}, called fast multivariate adaptive regression
splines (Fast MARS). In Fast MARS, a least squares fit model is
iteratively built by beginning with a single constant valued function
and adding two new basis functions at each iteration of the form
\begin{align*}
  B_{2j-1}(x) &= B_l(x) \bigl(x_i-x^{(p)}_i\bigr)_+, \\
  B_{2j}(x) &= B_k(x) \bigl(x_i-x^{(p)}_i\bigr)_- ,
\end{align*}
where \changed{$j \leq m$ is the iteration number, $m$ is the maximum
  number of underlying basis functions}{}, $1 \le p \le n$, and
$B_l(x)$, $B_k(x)$ are basis functions from the previous iteration,
 $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$ and
$w_- = (-w)_+$. After iteratively constructing a model, MARS then
iteratively removes basis functions that do not contribute to goodness
of fit. In effect, MARS creates a locally component-wise tensor
product approximation of the data. The overall computational
complexity of Fast MARS is $\mathcal{O}(n d m^3)$ \changed{}{where $m$
  is the maximum number of underlying basis functions}. This paper
uses an implementation of Fast MARS \cite{rudy2017pyearth} with $m =
200$ throughout all experiments.

\subsection{Support Vector Regressor}
\label{sec:svr}
Support vector machines are a common method used in machine learning
classification tasks that can be adapted for the purpose of regression
\cite{basak2007support}. How to build a support vector regressor (SVR)
is beyond the scope of this summary, but the resulting functional fit
$p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
 $$ p(x)  = \sum_{i=1}^{n}a_i K\bigl(x,x^{(i)}\bigr) + b ,$$
where $K$ is the selected kernel function, $a_i \in \mathbb{R}^n$, 
$b \in \mathbb{R}$ are coefficients to be solved for simultaneously.
The computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with
$m$ being determined by the minimization convergence criterion. This
paper uses the scikit-learn SVR \cite{scikit-learn} with a radial
basis function kernel.

\subsection{Multilayer Perceptron Regressor}
\label{sec:mlp}
The neural network is a well studied and widely used method for both
regression and classification tasks
\cite{hornik1989multilayer,rumelhart1988learning}. When using the
rectified linear unit (ReLU) activation function
\cite{dahl2013improving} and fitting the model with stochastic
gradient descent (SGD) or BFGS minimization techniques
\cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model
built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
\rightarrow \mathbb{R}^{j}$ defined by
 $$ l(u) = \bigl( u^t W_l + c_l \bigr)_+ ,$$
where $c_l \in \mathbb{R}^j$ and $W_l$ is the $i \times j$ weight
matrix for layer $l$. In this form, the multilayer perceptron (MLP)
produces a piecewise linear model of the input data. The computational
complexity of training a multilayer perceptron is $\mathcal{O}(n d
m)$, where $m$ is determined by the sizes of the layers of the network
and the stopping criterion of the minimization used for finding
weights. This paper uses an MLP built from Keras and Tensorflow to
perform regression \cite{tensorflow2015-whitepaper,chollet2015keras}
with ten hidden layers each having thirty two nodes (a total of
approximately ten thousand parameters), ReLU activation, and
\changed{one hundred}{five} thousand \changed{steps}{epochs} of SGD
for training. \changed{It should be noted that the choice of neural
  network architecture impacts approximation performance, but no
  architecture tuning is performed here}{}.

@@

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.4\textwidth,]{example_delaunay.pdf}
  \includegraphics[width=0.4\textwidth,]{example_not_delaunay.pdf}
  \caption{\color{red} On the left above is a depiction of a Delaunay
    triangulation over four points, notice that the circumball (shaded
    circle) for the left simplex does not contain the fourth point.
    On the right above, a non-Delaunay mesh is depicted. Notice that
    the circumball for the top simplex (shaded circle, clipped at
    bottom edge of the visual) contains the fourth point which
    violates the Delaunay condition for a simplex.
  \vspace{-.1cm}}
  \label{fig:delaunay}
\end{figure}
}{}

@@

The modified Shepard method used here (ShepMod) generates a continuous
approximation based on Euclidean distance and resembles a nearest
neighbor interpolant \cite{cover1967nearest}. This model is a type of
\textit{metric interpolation}, also called a Shepard method
\cite{gordon1978shepard,shepard1968two}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where \changed{$W_k(x^{(k)}) = 1$, $W_k(x) = \bigl(\bigl(r_k -
  \bigl\|x - x^{(k)}\bigr\|_2\bigr)_+ \big/ \bigl( r_k \bigl\|x -
  x^{(k)}\bigr\|_2 \bigr)\bigr)^2$ for $x \neq x^{(k)}$,}{} and $r_k
\in \mathbb{R}$ is the smallest radius such that at least $d+1$ other
points $x^{(j)}$, $j \not = k$, are inside the closed Euclidean ball
of radius $r_k$ about $x^{(k)}$. The computational complexity of
ShepMod is $\mathcal{O}(n^2d)$. This paper uses a Fortran 95
implementation of ShepMod derived from SHEPPACK
\cite{thacker2010algorithm}.

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.9\textwidth,]{example_shepard_with_weight.pdf}
  \caption{\color{red} On the left above is a depiction of the radius
    of influence for three chosen points of a collection in two
    dimensions using the modified Shepard criteria. On the right is
    the shape of the weight function $W_k$ for all data points, where
    $W_k(x) = 0$ when $\|x-x^{(k)}\|_2 \geq r_k$. It can also be seen
    that in the limit as $W_k$ tends towards infinity at $x_k$, the
    relative weight of $x_k$ tends to one.
  \vspace{-.1cm}}
  \label{fig:shepard}
\end{figure}
}{}

\subsection{Linear Shepard}
The linear Shepard method (LSHEP) is a blending function using local
linear interpolants, a special case of the general Shepard algorithm
\cite{thacker2010algorithm}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}
 {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where $W_k(x)$ is \changed{the same as for the modified Shepard
  method}{a locally supported weighting function} and $P_k(x)$ is a
local linear approximation to the data satisfying
$P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
Fortran 95 implementation of LSHEP in SHEPPACK
\cite{thacker2010algorithm}.

@@

The box spline model used here is \changed{an}{sparse grid}
interpolation technique built from overlapping box splines
\cite{de2013box}. The box splines serve as basis functions that are
shifted and scaled to have support over box shaped regions. The boxes
are constructed in a way to guarantee a covering of the domain
\cite{lux2018novel}. Given a set of box splines
$\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
in \cite{lux2018novel} and anchored at interior points
$\bigl\{x^{(i)}\bigr\}$,
 $$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
Note that the box splines always satisfy
$b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y)
\geq 0$. The computational complexity of interpolation via the box
spline model is $\mathcal{O}(n^2 d)$.

@@

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth,]{VM_Construction.pdf}
  \caption{\color{red} Above is a depiction of the Voronoi cell
    boundaries (dashed lines) about a set of interpolation points
    (dots) in two dimensions. In this example, the Voronoi model basis
    function about the center most point has nonzero weight in the
    shaded region and transitions from a value of one at the point to
    zero at the boundary of the twice expanded Voronoi cell (solid
    line).
  \vspace{-.1cm}}
  \label{fig:voronoi-basis}
\end{figure}
}{}

@@

\changed{
\begin{figure}
  \centering
  \includegraphics[width=0.8\textwidth,]{example_distribution_prediction.pdf}
  \caption{In this HPC I/O example, the general methodology for
    predicting a CDF and evaluating error can be seen\color{red},
    where M means $\times 10^6$\color{black}. The Delaunay method
    chose three source distributions (dotted lines) and assigned
    weights \color{red}\{.1, .3, .6\} \color{black} (top to bottom at
    \color{red}middle\color{black}). The weighted sum of the three
    known CDFs produces the predicted CDF (dashed line). The KS
    Statistic (\color{red}vertical line\color{black}) computed between
    the true CDF (solid line) and predicted CDF (dashed line) is 0.2
    for this example. For this example the KS test null hypothesis is
    rejected at $p$-value 0.01, however it is not rejected at
    $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction-example}
\end{figure}
}{}

@@

A hurdle when modeling function-valued outputs such as cumulative
distribution functions (CDFs) or probability density functions (PDFs)
is that certain properties must be maintained. It is necessary that a
PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. \changed{Instead}{However}, for a CDF
$F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
$F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and
nondecreasing. This work utilizes the fact that a convex combination
of CDFs (or PDFs) results in a valid CDF (or PDF). Given $G(x) =
\sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$
is a valid CDF, $G$ must also be a valid CDF. A demonstration of how
this is applied can be seen in Figure \ref{fig:prediction-example}. 
\changed{In this example the KS test null hypothesis is rejected at $p$-value
0.01, however it is not rejected at $p$-value 0.001.}{}

@@

The two-sample KS test is a useful nonparametric test for comparing
two \changed{empirical}{} CDFs while only assuming stationarity,
finite mean, and finite variance. The null hypothesis (that two
\changed{empirical}{} CDFs come from the same underlying distribution)
is rejected at level $p \in [0,1]$ when
 $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)}
\sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$ with distribution sample
sizes $n_1,n_2 \in \mathcal{N}$. For all applications of the KS test
presented in this work $n_1 = n_2$. An example of \changed{the process
  of generating predicted distributions from known distributions and
  the subsequent}{the round-trip prediction methodology from known and
  predicted distributions to the} calculation of error can be seen in
Figure \ref{fig:prediction-example}. A brief listing of relevant
statistical terms used throughout this work is provided in the
Appendix (Section \ref{sec:appendix}).

@@

This section presents the theoretical results bounding the error of
(piecewise) linear interpolation. The error analysis relies on linear
interpolation for three reasons: (1) second order results can be
obtained utilizing a Lipschitz constant on the gradient of a function,
rather than standard Lipschitz bounds; (2) the results directly apply
to Delaunay interpolation; and (3) multiple other interpolants in this
paper compute predictions as convex combinations of observed function
values, which may allow for straightforward extensions of this error
bound.

\begin{plainlemma}
  \label{lemma:1}
  Let $S \subset \mathbb{R}^d$ be open and convex, $f: S \rightarrow
  \mathbb{R}$, and let $\nabla f(x)$ be $\gamma$-Lipschitz continuous
  in the $2$-norm. Then for all $x,y \in S$
  $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
\end{plainlemma}

\begin{proofdot}
  Consider the function $g(t) = f \big((1-t) x + t y \big)$, $0 \leq t
  \leq 1$, whose derivative $g'(t) = \big\langle \nabla f \big((1-t) x
  + t y \big), y - x \big\rangle$ is the directional derivative of $f$
  in the direction $(y - x).$
  \begin{align*}
    \big|f(y) - f(x) - \langle &\nabla f(x), y - x \rangle \big|
        = \big|g(1) - g(0) - g'(0) \big| \\
       &= \bigg| \int_0^1 g'(t) - g'(0)\ dt \bigg| \leq \int_0^1 \big|g'(t) - g'(0)\big|\ dt \\
       &= \int_0^1 \bigg| \big \langle \nabla f\big((1-t)x + ty\big) - \nabla f(x), y - x \big \rangle \bigg|\ dt \\
       &\leq \int_0^1 \big \| \nabla f\big((1-t)x + ty\big) - \nabla f(x) \big \|_2\ \| y - x \|_2\ dt \\
       &\leq \int_0^1 \big ( \gamma\ \|y-x\|_2 \big) \ \big( \|y-x\|_2 \big) t\ dt = \frac{\gamma \|y - x\|_2^2}{2}.
  \end{align*}
  \qed
\end{proofdot}

\hfill

>>>>

% Insert here the call for the packages your document requires
\usepackage{graphicx}
\usepackage{tikz}     % For adding axes to plots
\usetikzlibrary{positioning}
\usepackage{graphicx} % For including PDFs
\usepackage{amsmath}  % assumes amsmath package installed
\usepackage{amssymb}  % assumes amsmath package installed
\usepackage{booktabs} % For formal tables
\usepackage{tabularx} % For tables with nice column widths.
\usepackage[colorlinks,allcolors=blue]{hyperref}
\usepackage{float}    % For controlling the placement of tables.

@@

\title{Interpolation of Sparse High-Dimensional Data\thanks
  {This work was supported by the National Science Foundation Grant CNS-1565314.}
}
% \subtitle{Do you have a subtitle?\\ If so, write it here}
%\titlerunning{Short form of title}        % if too long for running head

@@

\date{Received: date / Accepted: date}
% The correct dates will be entered by the editor

\maketitle

@@

Regression and interpolation have a considerable theoretical base in
one dimension \cite{cheney2009course}. Splines in particular are well
understood as an interpolation technique in one dimension
\cite{de1978practical}, particularly B-splines. Tensor products of
B-splines \cite{unther1996interpolating} or other basis functions have
an unfortunate exponential scaling with increasing
dimension. Exponential scaling prohibits tensor products from being
reasonably applied beyond three-dimensional data. In order to address
this dimensional scaling challenge, C. de Boor and others proposed box
splines \cite{de2013box}, of which one of the approximation techniques
in this work is composed \cite{lux2018novel}.

The theoretical foundation of low dimensional interpolation allows the
construction of strong error bounds that are absent from high
dimensional problems. This work extends some known results regarding
the secant method \cite{dennis1996numerical} to construct an
interpolation error bound for problems of arbitrary dimension. These
error bounds are useful, considering the same cannot be said for
regression algorithms in general. The maximum complexity of an
interpolant is bounded by the amount of data available, while the
maximum complexity of a regressor is bounded by both the amount of
data and the chosen parametric form. For this reason, generic uniform
bounds are largely unobtainable for regression techniques on arbitrary
approximation problems, even when the approximation domain is bounded.

@@

In this work, multivariate interpolation is defined given a closed
convex subset $Y$ of a metrizable topological vector space with metric
$s$, some function $f:\mathbb{R}^d \rightarrow Y$ and a set of points
$X = \bigl\{x^{(1)}$, $\ldots$, $x^{(n)}\bigr\} \subset \mathbb{R}^d$,
along with associated function values $f\bigl(x^{(i)}\bigr)$. The goal
is to construct an approximation $\hat f: \mathbb{R}^d \rightarrow Y$
such that $\hat f\bigl(x^{(i)}\bigr) = f\bigl(x^{(i)}\bigr)$ for all
$i = 1$, $\ldots$, $n$. It is often the case that the form of the true
underlying function $f$ is unknown, however it is still desirable to
construct an approximation $\hat f$ with small approximation error at
$y \notin X$. The two metric spaces that will be discussed in this
work are the real numbers with metric $s(x,y) = |x-y|$, and the set of
cumulative distribution functions (CDFs) with the Kolmogorov-Smirnov
(KS) statistic \cite{lilliefors1967kolmogorov} as metric.

@@

The most challenging problem when scaling in dimension is that the
number of possible interactions between dimensions grows
exponentially. Quantifying all possible interactions becomes
intractable, and hence beyond three-dimensional data mostly linear
models are used. That is not to say nonlinear models are absent, but
nonlinearities are often either preconceived or model pairwise
interactions between dimensions at most. Even globally nonlinear
approximations such as neural networks are constructed from
compositions of summed low-interaction functions
\cite{clevert2015fast}.

Provided the theoretical and practical motivations for exploring
interpolants, this work aims to study the empirical performance
differences between a set of scalable (moderately) interpolation
techniques and a set of common regression techniques. These techniques
are applied to a collection of moderately high dimensional problems
($5 \le d \le 30$) and the empirical results are discussed.

The remainder of this paper is organized as follows. Sections
\ref{sec:regression} and \ref{sec:interpolation} present the
multivariate models. Section \ref{sec:error} gives the error measuring
methodology that is used for collecting empirical results. Section
\ref{sec:theory} contains the theoretical error bounds for
interpolation. Section \ref{sec:data} presents the data sets for
empirical approximation analysis and the approximation results for all
models on these data sets. Section \ref{sec:discussion} analyzes and
discusses the results, their implications, and their
limitations. Finally, Section \ref{sec:conclusion} concludes.

@@

\subsection{Multivariate Adaptive Regression Splines}
\label{sec:mars}
This approximation was introduced in \cite{friedman1991multivariate}
and subsequently improved to its current version in
\cite{stanford1993fast}, called fast multivariate adaptive regression
splines (Fast MARS). In Fast MARS, a least squares fit model is
iteratively built by beginning with a single constant valued function
and adding two new basis functions at each iteration of the form
\begin{align*}
  B_{2j-1}(x) &= B_l(x) \bigl(x_i-x^{(p)}_i\bigr)_+, \\
  B_{2j}(x) &= B_k(x) \bigl(x_i-x^{(p)}_i\bigr)_- ,
\end{align*}
where $j$ is the iteration number, $1 \le p \le n$, and $B_l(x)$,
$B_k(x)$ are basis functions from the previous iteration,
 $$w_+ = \begin{cases} w, & w \geq 0 \\ 0, & w < 0 \end{cases},$$
and $w_- = (-w)_+$. After iteratively constructing a model, MARS then
iteratively removes basis functions that do not contribute to goodness
of fit. In effect, MARS creates a locally component-wise tensor
product approximation of the data. The overall computational
complexity of Fast MARS is $\mathcal{O}(n d m^3)$ where $m$ is the
maximum number of underlying basis functions. This paper uses an
implementation of Fast MARS \cite{rudy2017pyearth} with $m = 200$
throughout all experiments.

\subsection{Support Vector Regressor}
\label{sec:svr}
Support vector machines are a common method used in machine learning
classification tasks that can be adapted for the purpose of regression
\cite{basak2007support}. How to build a support vector regressor (SVR)
is beyond the scope of this summary, but the resulting functional fit
$p : \mathbb{R}^d \rightarrow \mathbb{R}$ has the form
 $$ p(x)  = \sum_{i=1}^{n}a_i K\bigl(x,x^{(i)}\bigr) + b ,$$
where $K$ is the selected kernel function, $a_i \in \mathbb{R}^n$, 
$b \in \mathbb{R}$ are coefficients to be solved for simultaneously.
The computational complexity of the SVR is $\mathcal{O}(n^2dm)$, with
$m$ being determined by the minimization convergence criterion. This
paper uses the scikit-learn SVR \cite{scikit-learn} with a polynomial
kernel function.

\subsection{Multilayer Perceptron Regressor}
\label{sec:mlp}
The neural network is a well studied and widely used method for both
regression and classification tasks
\cite{rumelhart1988learning,hornik1989multilayer}. When using the
rectified linear unit (ReLU) activation function
\cite{dahl2013improving} and fitting the model with stochastic
gradient descent (SGD) or BFGS minimization techniques
\cite{goh2017why,moller1993scaled,robbins1951stochastic}, the model
built by a multilayer perceptron uses layers $l : \mathbb{R}^{i}
\rightarrow \mathbb{R}^{j}$ defined by
 $$ l(u) = \bigl( u^t W_l + c_l \bigr)_+ ,$$
where $c_l \in \mathbb{R}^j$ and $W_l$ is the $i \times j$ weight
matrix for layer $l$. In this form, the multilayer perceptron (MLP)
produces a piecewise linear model of the input data. The computational
complexity of training a multilayer perceptron is $\mathcal{O}(n d
m)$, where $m$ is determined by the sizes of the layers of the network
and the stopping criterion of the minimization used for finding
weights. This paper uses an MLP built from Keras and Tensorflow to
perform regression \cite{tensorflow2015-whitepaper,chollet2015keras}
with ten hidden layers each having thirty two nodes (a total of
approximately ten thousand parameters), ReLU activation, and five
thousand epochs of SGD for training.

@@

@@

The modified Shepard method used here (ShepMod) generates a continuous
approximation based on Euclidean distance and resembles a nearest
neighbor interpolant \cite{cover1967nearest}. This model is a type of
\textit{metric interpolation}, also called a Shepard method
\cite{shepard1968two,gordon1978shepard}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)f\bigl(x^{(k)}\bigr)}
     {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where $W_k(x) = \bigl(r_k - \bigl\|x - x^{(k)}\bigr\|\bigr)_+^2 \big/
\bigl\|x - x^{(k)}\bigr\|_2^2$ and $r_k \in \mathbb{R}$ is the
smallest radius such that at least $d+1$ other points $x^{(j)}$, $j
\not = k$, are inside the closed Euclidean ball of radius $r_k$ about
$x^{(k)}$. The computational complexity of ShepMod is
$\mathcal{O}(n^2d)$. This paper uses a Fortran 95 implementation of
ShepMod derived from SHEPPACK \cite{thacker2010algorithm}.

\subsection{Linear Shepard}
The linear Shepard method (LSHEP) is a blending function using local
linear interpolants, a special case of the general Shepard algorithm
\cite{thacker2010algorithm}. The interpolant has the form
 $$ p(x) = \frac{\sum\limits_{k=1}^{n}W_k(x)P_k(x)}
 {\sum\limits_{k=1}^{n}W_k(x)} ,$$
where $W_k(x)$ is a locally supported weighting function and $P_k(x)$
is a local linear approximation to the data satisfying
$P_k\bigl(x^{(k)}\bigr) = f\bigl(x^{(x)}\bigr)$. The computational
complexity of LSHEP is $\mathcal{O}(n^2d^3)$. This paper uses the
Fortran 95 implementation of LSHEP in SHEPPACK
\cite{thacker2010algorithm}.

@@

The box spline model used here is an interpolation technique built
from overlapping box splines \cite{de2013box}. The box splines serve
as basis functions that are shifted and scaled to have support over
box shaped regions. The boxes are constructed in a way to guarantee a
covering of the domain \cite{lux2018novel}. Given a set of box splines
$\bigl\{b^{x^{(i)}}\bigr\}$ with the iterative box properties outlined
in \cite{lux2018novel} and anchored at interior points
$\bigl\{x^{(i)}\bigr\}$,
 $$ \hat f(y) = \frac{\sum\limits_{i=1}^n b^{x^{(i)}}(y) f\bigl(x^{(i)}\bigr)}{\sum\limits_{i=1}^n b^{x^{(i)}}(y)}. $$
Note that the box splines always satisfy
$b^{x^{(i)}}\bigl(x^{(j)}\bigr) = \delta_{ij}$ and $b^{x^{(i)}}(y)
\geq 0$. The computational complexity of interpolation via the box
spline model is $\mathcal{O}(n^2 d)$.

@@

@@

\begin{figure}[htb]
  \centering
  \begin{tikzpicture}
    \node (img)  {\includegraphics[width=0.75\textwidth,]{Fig1.pdf}};
    \node[below=of img, node distance=1cm, yshift=1cm] {I/O Throughput};
    \node[left=of img, node distance=0cm, rotate=90, anchor=center,yshift=-0.7cm] {CDF Value};
  \end{tikzpicture}
  \vspace{-0.3cm}
  \caption{In this HPC I/O example, the general methodology for
    predicting a CDF and evaluating error can be seen. The Delaunay
    method chose three source distributions (dotted lines) and
    assigned weights \{.3, .4, .3\} (top to bottom at arrow). The
    weighted sum of the three known CDFs produces the predicted CDF
    (dashed line). The KS Statistic (arrow) computed between the true
    CDF (solid line) and predicted CDF (dashed line) is 0.2 for this
    example. The KS test null hypothesis is rejected at $p$-value
    0.01, however it is not rejected at $p$-value 0.001.
  \vspace{-.1cm}}
  \label{fig:prediction-example}
\end{figure}

@@

A hurdle when modeling function-valued outputs such as cumulative
distribution functions (CDFs) or probability density functions (PDFs)
is that certain properties must be maintained. It is necessary that a
PDF $f: \mathbb{R} \rightarrow \mathbb{R}$ have the properties $f(x)
\geq 0$ and $\int_{-\infty}^{\infty}f(x)dx = 1$. However, for a CDF
$F: \mathbb{R} \rightarrow \mathbb{R}$ the properties are
$F(x) \in [0,1]$ and $F(x)$ is absolutely continuous and
nondecreasing. This work utilizes the fact that a convex combination
of CDFs (or PDFs) results in a valid CDF (or PDF). Given $G(x) =
\sum_{i}w_i F_i(x)$, $\sum_{i} w_i = 1$, $w_i \geq 0$, and each $F_i$
is a valid CDF, $G$ must also be a valid CDF. A demonstration of how
this is applied can be seen in Figure \ref{fig:prediction-example}.

@@

The two-sample KS test is a useful nonparametric test for comparing
two CDFs while only assuming stationarity, finite mean, and finite
variance. The null hypothesis (that two CDFs come from the same
underlying distribution) is rejected at level $p \in [0,1]$ when
 $$ KS > \sqrt{-\frac{1}{2}\ln\biggl(\frac{p}{2}\biggr)} \sqrt{\frac{1}{n_1} + \frac{1}{n_2}}, $$
with distribution sample sizes $n_1,n_2 \in \mathcal{N}$. For all
applications of the KS test presented in this work $n_1 = n_2$. An
example of the round-trip prediction methodology from known and
predicted distributions to the calculation of error can be seen in
Figure \ref{fig:prediction-example}.

@@

This section presents the theoretical results bounding the error of
(piecewise) linear interpolation. The error analysis relies on linear
interpolation for three reasons: (1) second order results can be
obtained utilizing a Lipschitz constant on the gradient of a function,
rather than standard Lipschitz bounds; (2) the results directly apply
to Delaunay interpolation; and (3) multiple other interpolants in this
paper compute predictions as convex combinations of observed function
values, which may allow for straight forward extensions of this error
bound.

\begin{plainlemma}
  \label{lemma:1}
  Let $S \subset \mathbb{R}^d$ be open and convex, $f: \mathbb{R}^d
  \rightarrow \mathbb{R}$, and $\nabla f \in
  Lip_{(\gamma,\|\cdot\|_2)}(S)$, the set of $\gamma$-Lipschitz
  continuous functions in the $2$-norm. Then for all $x,y \in S$
  $$\big|f(y) - f(x) - \langle \nabla f(x), y - x \rangle \big| \leq \frac{\gamma \|y - x\|_2^2}{2}.$$
\end{plainlemma}
